---
title: "lab_02"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
library(stringr)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```
```{r}
glimpse(umd_courses)
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? 

**A1.** Atmospheric and Oceanic Science offered the most amount of classes with Climate in the title. 

```{r}
climate_courses <- umd_courses |>
  filter(str_detect(title, "Climate")) |>
  filter(seats >= 1) 
```


```{r}
climate_courses_summary <- climate_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    max_seats = max(seats, na.rm = TRUE),
    min_seats = min(seats, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(max_seats), desc(min_seats))
```


**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? 

**A2.** Art History because I am interested to see how their classes approach the discussion of climate through art, or documenting climate change through art. 

**Q3** Which departments offered the most courses in the Fall 2024 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3.** Business and Management offered the most amount of courses, 174, in fall 2024. The departments with at least one syllabus posted are listed under the column, syllabi_posted. The top five departments are different in how many syllabi they've posted for each course, some, like Business and Management, have a high amount of courses with a corresponding amount of posted syllabi while others, like History, have 22 posted syllabi to 116 offered courses. 

```{r}
fall_2024_courses <- umd_courses |>
  filter(term == 202408) |>
  group_by(department) |>
  summarise(
    num_courses_depart = n(),
    syllabi_posted = sum(syllabus_count >= 1, na.rm = TRUE), 
  )|>
   filter(syllabi_posted >= 1) |>  
arrange(desc(num_courses_depart), desc(syllabi_posted))

head(fall_2024_courses)
```
**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time.

**A4.** The data reveals a pattern where the amount of money spent by Moore's committee drastically increases , reaching over $3 million by October 2022, over 1000% more than previous year. 



```{r}
glimpse(maryland_expenses)
head(maryland_expenses)
```


```{r}
# Convert expenditure_date to date and create month column
maryland_expenses <- maryland_expenses |>
mutate(expenditure_date = mdy(expenditure_date)) 
```
```{r}
maryland_expenses |>
mutate(month = floor_date(expenditure_date, "month")) |>
 group_by(month) |>
  summarise(total_amount = sum(amount, na.rm = TRUE)) |>
  arrange(desc(total_amount))
```
```{r}
wes_moore_expenses <- maryland_expenses|>
filter(str_detect(committee_name, "Moore")) |>
mutate(month = floor_date(expenditure_date, "month")) |>
  group_by(month) |>
  summarise(total_amount = sum(amount, na.rm = TRUE)) |>
  arrange(desc(total_amount))

wes_moore_expenses
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe). **A5** In-State : 7,918,989	and Out-state: 3,843,620

```{r}
wes_moore_expenses <- maryland_expenses |>
  filter(str_detect(committee_name, "Moore")) |>
 mutate(
   spending_location = case_when( 
        str_detect(address, "Maryland") ~ "In-State",
        TRUE ~ "Out of state"
      ))

```

```{r}
spending_summary <- wes_moore_expenses |>
  group_by(spending_location) |>
  summarise(
    total_spent = sum(amount, na.rm = TRUE),
  )

spending_summary
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. **A6.**

Q6: identify the different types of expense methods and how many times cash was used 

```{r}
maryland_expenses |>
  group_by(expense_method)|>
 summarise(
    times_used= n()
 ) |>
arrange(desc(times_used))
```
Filter for when the expense method is NA. (I used AI for this, my prompt was "i want to create an internal table with expense methods marked NA" with the code I'd written that wasn't working. What was missing was "is.na," which tells R to filter anything in expense_method that is labled NA)

```{r}
maryland_expenses |>
  group_by(expense_method) |>
  filter(is.na(expense_method)) 
```
