---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the \~ is equivalent of /Users/[youruser] on a Mac).

IMPORTANT or you WILL go insane: - remember quotations "" '' around variables

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)

```

```{r}
filter()
```

\*\* creates a new data set with the filtered information/rows

```{r}
select()
```

\*\*creates a data set for only the selected columns

```{r}
#umd_courses |>
  filter(str_detect(title, "Shakespeare"))
```

# Creating Data Frame

NOTE: `readr` anticipates some date formatting and can automatically handle many of these issues (indeed it uses lubridate under the hood). When you are importing a CSV file, be sure to use `read_csv`, not `read.csv`.

```{r}
maryland_expenses <- read_csv("data/maryland_expenses.csv")

head(maryland_expenses)
```

-   head() shows only first 6 columns

# Filters

\*\* When it comes to filters, OR; `|` is additive; it adds more possibilities to the search criteria

AND; `&` is restrictive; it imposes more conditions, limiting the results to only those that meet all criteria.

str_detect --\> string detect (reads text) \*used inside filter and give it a column (title) to search and what you are looking for (shakespeare)

str_like: uses the wildcard search operator % to do partial matching for words

# Dates

Lubricate --\> a tidyverse library created to deal with dates [YYYY-MM-DD]

-   floor date: used to add up the total amount spent in each month in our Maryland expenses data, for example.

```{r}
#maryland_expenses <- maryland_expenses |> mutate(expenditure_date=mdy(expenditure_date))

#head(maryland_expenses)
```

summarise(total_amount= sum(amount)): sums up

\`str_to_upper --\> a mix of styles: lower-case and upper-case names like "Anedot" and "ANEDOT", R will think are different

every group_by takes the argument of the column

count= N() --\> counts the amount of cells in

IQR
```{r}
wh_salaries_clean |>
  summarise(
    Q1 = quantile(salary, 0.25, na.rm = TRUE),
    Q3 = quantile(salary, 0.75, na.rm = TRUE),
    IQR = IQR(salary, na.rm = TRUE)
  ) |>
  mutate(
    lower_bound = Q1 - 1.5 * IQR,
    upper_bound = Q3 + 1.5 * IQR
  )
```
Then filter salaries outside that range:

```{r}
wh_salaries_clean |>
  filter(salary < lower_bound | salary > upper_bound)
```

