---
title: "lab_05"
author: "Derek Willis"
date: "2025-01-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   Tabula

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse, plus any other packages you will need to clean data and work with dates.
library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
```

## Get Our PDF

We'll be working with the [911 overdose calls from Baltimore County](https://drive.google.com/file/d/1qkYuojGF_6WKFr5aNQxmewDzcKyOiJFr/view?usp=share_link). You'll want to download it to a place you'll remember (like your Downloads folder, or the labs folder in your repository). The goal is to extract the tables within it, export that to a CSV file, load it into RStudio and ask some questions.

## Extract Data from PDF Using Tabula

Start Tabula, then go to <http://127.0.0.1:8080/> in your browser. Click the "Browse" button and find the PDF file and click "open", and then click the "Import button" in Tabula. This will take a few seconds or longer.

This PDF has a single table spread over multiple pages to extract. We're going to make a single dataframe from this table, exporting it to a CSV file that you will load into R. In Tabula, highlight the table and click the "Preview & Export Extracted Data" button. You may want to play with including or excluding the column headers - YOU SHOULD HAVE FIVE COLUMNS OF DATA.

Save the CSV (it should be called `tabula-Baltimore County; Carey, Samantha log OD.csv` by default) to your lab_05/data folder.

From there, you will need to read in the data, and add or fix headers if necessary. You can choose to include the headers from the PDF in your exported CSV files OR to exclude them and add them when importing. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)).

## Load and clean up the data in R

You will need to read in and clean up the data so that it can be used for analysis. By "clean" I mean the column headers should not contain spaces and they should have meaningful names, not "x1" or something similar. How you do that is up to you, but you can use select() with or without the minus sign to include or exclude certain columns. You also can use the `rename` function to, well, rename columns. Importantly, you'll need to ensure that any columns containing a date actually have a date datatype. Our friend `lubridate` can help with this.

```{r} 
balt_overdose <- read_csv("tabula-Baltimore County; Carey, Samantha log OD.csv") |>
  clean_names()
```


```{r}
cleaned_balt_overdose <- balt_overdose |>
   mutate(
    date = mdy(date),
    location = location |>
      str_replace_all("&", "AND") |>      
      str_remove("\\(.*?\\)") |>           
      str_remove("\\*.*?\\*") |>           
      str_remove(";.*") |>                
      str_squish()                        
  )
cleaned_balt_overdose
                               
```

## Answer questions

Q1. Write code to generate the number of calls that occurred on each date. Which date in 2022 had the most overdose calls, and how many? Look at the total number of rows in your result and explore the range of dates - based on your result, do you believe there are any days with no overdose calls at all? Explain why or why not.

A1. July 14th 2022 had the most overdose calls at 23, so did October 4th. No I do not believe there are dates with no overdose calls because I arranged the dates by descending order and looking through all dates are accounted for and have a positive count. 

```{r}
cleaned_balt_overdose |>
  group_by(date) |>
  summarise(count = n()) |>
 arrange(desc(count)) 
```

Q2. You want to understand if there's a pattern in the day of the week that overdose calls are made. Add a column to your dataframe that displays what day of the week each date represents. You should search for how to do that using lubridate. Then write code to calculate the number of calls for each day of the week, and add a column to that result that calculates the percentage of all calls that occurred on each day of the week (so you want a dataframe with the day of the week, total number of calls and the percentage of calls on that day out of the total number of all calls). Describe your findings to me.

A2. Based on my findings, Thursday has the fewest overdose calls and the lowest percentage change per 1000. Though from Thursday (127.9) to Saturday (155.2), there’s about a 21% increase in overdose call rates. This suggests a clear connection that the overdose calls increase as the weekend rolls around. Saturday (638 calls) is the peak, followed by Sunday and Friday. This isn't surprising and aligns with human and public health patterns where higher substance use or risk-taking tends to occur more frequently on weekends.This also means that per 1000 residents, about 155 overdose calls happen on Saturdays vs 128 on Thursdays. 

```{r}
cleaned_balt_overdose |>
mutate(day_of_week = wday(date, label = TRUE)) |>
group_by(day_of_week) |>
summarise(count_by_day = n()) |>
mutate(percent_of_total = (count_by_day / sum(count_by_day)) * 1000) |>
arrange(desc(count_by_day)) 
```

Q3. Now let's look at locations. Which ones have the most calls? How would you describe them (feel free to search for more information on them)? Is there anything about the structure of the original data that might make you less confident in the counts by location or date?

A3. The location with the most overdose calls, 36, is 4540 Silver Spring Rd Nottingham, MD 21236, a 3 bedroom, 2 bath house that is currently occupied. The second is Royal Farms, one of the four grocery/gas stations in the top ten locations for overdose calls. In the original dataset, the location column has entries such as FOOD LION; 7514 NORTH POINT RD, which could be a problem when trying to standardize because its inconsistent with the formatting for other similar locations. Additionally, there is inconsistent punctuation on some address. The first code block I wrote should've cleaned up most of these inconsistencies enough for me to feel confident by the count_by_location output. 

```{r}
cleaned_balt_overdose |>
group_by(location) |>
summarise(count_by_location = n()) |>
arrange(desc(count_by_location)) 
```

Q4. What's the best story idea or question you've seen as a result of the work you've done in this lab?
 
PC 02 and PC 06 are police precincts in Woodlawn and Towson, and they have the third and fourth amount of overdose calls respectively. This indicates that some overdoses occured in police custody. This raises critical questions about the circumstances of these overdoses, including whether affected individuals were given the necessary proper care and if these locations had fatalities and why.
